STJ2012 analysis

### 2016-08-01 analysis/STJ2012_analysis.Rmd
add individual spp barplots

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, fig.height=4, fig.width=8}
# Compute summary statistics
stats <- function(phy) {
  return(
    data.frame(rangeotus=paste0(range(taxa_sums(phy)), collapse=" - "),
               nsingles=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
               rangerps=paste0(range(sample_sums(phy)), collapse=" - "),
               amrps=paste(as.integer(mean(sample_sums(phy))), 
                           as.integer(sd(sample_sums(phy))), sep=" ± "),
               gmrps=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                           as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "))
  )
}
stats97 <- stats(phy97)
stats100 <- stats(phy100)

@@

## Filter dataset
#### Filter samples by minimum count
```{r filter_samples}
# Set threshold number of reads
sn <- 10
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)

@@

#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]  
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)  
phy100.f <- prune_taxa(names(taxa100), phy100.f)  
```

@@

## Beta diversity
Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group. Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test. This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
```{r betadisper, cache=T, fig.width=4, fig.height=4}
# Calculate distances to species centroids for each sample in principal coordinate space
sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
                    group=data.frame(sample_data(phy97.f.t))$Species, type="centroid",
                    bias.adjust=TRUE)
# Calculate each species' mean distance to centroid and standard error
sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
                      FUN=function(x) sd(x)/sqrt(length(x)))$x
# Determine order of decreasing mean dispersion
sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
# Update betadisper object with species in decreasing order of mean dispersion
sambd <- betadisper(d=vegdist(t(otu_table(phy97.f.t)), method="bray"), 
                    group=factor(data.frame(sample_data(phy97.f.t))$Species, levels=as.character(sambdsumm.ord$group)), type="centroid", bias.adjust=TRUE)
# Use permutations to perform pairwise comparisons of group mean dispersions
sampt <- permutest(sambd, pairwise=T)
saml <- multcompLetters(sampt$pairwise$permuted)

# Figure: Distance to centroids
par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
with(sambdsumm.ord, {
  plot(mean, type="n", ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=saml$Letters[as.character(group)], cex=0.5)
})
```

```{r strigosa}
dstr97 <- subset_samples(phy97.f.p, Species=="strigosa")
plot_bar(dstr97, x="Sample", y="Abundance", fill="OTU")
plot_bar(dstr97, x="Sample", y="Abundance", fill="Subtype")
# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
# Also try 100% OTUs -- must filter more first...
dstr100.f <- subset_samples(phy100.f, Species=="strigosa")
dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
#dstr100.f.p.f <- prune_taxa(rownames(otu_table(dstr100.f.p))[which(rowSums(otu_table(dstr100.f.p)) > 0.01)],
#                            dstr100.f.p)
#  
#plot_bar(dstr100.f.p.f, x="Sample", y="Abundance", fill="OTU")
#plot_bar(dstr100, x="Sample", y="Abundance", fill="Subtype")
#sample_names(dstr100.f)
#subset_samples(dstr100, Sample=="7")
#otu_table(dstr100.f)
#dstr100.f.p <- transform_sample_counts(dstr100.f, function(x) x/sum(x))
#colSums(otu_table(dstr100.f.p)[rowSums(otu_table(dstr100.f.p)) > 0.1, ])

@@

```{r otu_barplots_custom}
# 100%-OTUs barplot 
par(mar=c(2,3,4,1))
samdat <- data.frame(sample_data(dstr100.f.p))
typerelabund <- as.matrix(otu_table(dstr100.f.p)[order(data.frame(tax_table(dstr100.f.p))$Subtype), 
                                                 rownames(samdat)])

# Get info for plotting OTU names and blast hits on top of barplot
blasthits <- as.character(data.frame(tax_table(dstr100.f.p))[order(data.frame(tax_table(dstr100.f.p))$Subtype)
                                                             , "Subtype"])
names <- as.character(rownames(data.frame(tax_table(dstr100.f.p)))[order(data.frame(tax_table(dstr100.f.p))$Subtype)])
divides <- apply(typerelabund, 2, cumsum)
heights <- diff(divides)
heights[heights < 0.04] <- NA

# Plot barplot and OTU names and blast hits
bars <- barplot(typerelabund, col=rainbow(941), border=NA, las=1, xlab="Sample", ylab="", cex.axis=0.75, cex.names=0.75)
mtext(side=2, text = "Relative Abundance", line=2)
for (i in 1:length(bars)) {
  text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
       divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
       labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
       cex=0.6)
}

mtext(side=3, "B.) 100% OTUs", xpd=T, adj=0, line=2.5, font=2)

>>>>

#### Descriptive statistics and distribution of OTU counts
```{r phy_stats, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats <- function(phy) {
  return(
    data.frame(rangeotus=paste0(range(taxa_sums(phy)), collapse=" - "),
               nsingles=length(taxa_sums(phy)[taxa_sums(phy) <= 1]),
               rangerps=paste0(range(sample_sums(phy)), collapse=" - "),
               amrps=paste(as.integer(mean(sample_sums(phy))), 
                           as.integer(sd(sample_sums(phy))), sep=" ± "),
               gmrps=paste(as.integer(exp(mean(log(sample_sums(phy))))),
                           as.integer(exp(sd(log(sample_sums(phy))))), sep=" ± "))
  )
}
stats97 <- stats(phy97)
stats100 <- stats(phy100)

@@

## Filter dataset
#### Filter samples by minimum count
```{r filter_samples, cache=T}
# Set threshold number of reads
sn <- 10
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97)>=sn, phy97)
phy100.f <- prune_samples(sample_sums(phy100)>=sn, phy100)

@@

#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]  
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)  
phy100.f <- prune_taxa(names(taxa100), phy100.f)

# Refilter samples in case removal of OTUs caused any samples to drop below threshold
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
```

@@

## Beta diversity
Beta diversity is evaluated here as the multivariate dispersion of samples within a coral species group. Principal coordinate analysis on Bray-Curtis dissimilarities is used to calculate average distance-to-centroid values for each species group, which are then compared statistically by a permutation test. This analysis is implemented using betadisper() in the vegan package, based on [Anderson (2006)](dx.doi.org/10.1111/j.1541-0420.2005.00440.x).
```{r betadisper, cache=T, fig.width=8, fig.height=4}
betad <- function(phy) {
  # Calculate distances to species centroids for each sample in principal coordinate space
  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
                      group=data.frame(sample_data(phy))$Species, type="centroid",
                      bias.adjust=TRUE)
  # Calculate each species' mean distance to centroid and standard error
  sambdsumm <- aggregate(data.frame(mean=sambd$distances), by=list(group=sambd$group), FUN=mean)
  sambdsumm$se <- aggregate(sambd$distances, by=list(group=sambd$group), 
                            FUN=function(x) sd(x)/sqrt(length(x)))$x
  # Determine order of decreasing mean dispersion
  sambdsumm.ord <- sambdsumm[order(sambdsumm$mean, decreasing=T), ]
  # Update betadisper object with species in decreasing order of mean dispersion
  sambd <- betadisper(d=vegdist(t(otu_table(phy)), method="bray"), 
                      group=factor(data.frame(sample_data(phy))$Species,
                                   levels=as.character(sambdsumm.ord$group)), 
                      type="centroid", bias.adjust=TRUE)
  # Use permutations to perform pairwise comparisons of group mean dispersions
  sampt <- permutest(sambd, pairwise=T)
  saml <- multcompLetters(sampt$pairwise$permuted)
  # Return results as a list
  return(list(sambdsumm.ord=sambdsumm.ord, sambdsumm=sambdsumm, saml=saml))
}

betad97 <- betad(phy97.f.t)
betad100 <- betad(phy100.f.t)

# Figure: Distance to centroids
par(mfrow=c(1,2), mar=c(3.5,4,2,1), lwd=1)
with(betad97$sambdsumm.ord, {
  plot(mean, type="n", main="97% OTUs",
       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(betad97$sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=betad97$saml$Letters[as.character(group)], cex=0.5)
})
with(betad100$sambdsumm.ord, {
  plot(mean, type="n", main="100% OTUs",
       ylim=c(0, 0.75), ylab="Distance to centroid", xaxt="n", xlab="")
  arrows(1:10, mean - se, 1:10, mean + se, length=0.05, angle=90, code=3)
  points(1:10, mean, cex=1, pch=21, bg="white")
  text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75, 
       labels=levels(SAM$Species)[order(betad100$sambdsumm$mean, decreasing=T)])
  text(1:10, mean + se + 0.03, labels=betad100$saml$Letters[as.character(group)], cex=0.5)
})
```

```{r otu_barplots_custom, fig.width=8, fig.height=8}
otubarplot <- function(phy, main) {
  samdat <- data.frame(sample_data(phy))
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), 
                                                 rownames(samdat)])
  typerelabund <- typerelabund[, rownames(sample_data(phy)[with(sample_data(phy), 
                                                                order(Shore, Site, InputFileName))])]
  # Get info for plotting OTU names and blast hits on top of barplot
  blasthits <- as.character(data.frame(tax_table(phy))[order(data.frame(tax_table(phy))$Subtype), "Subtype"])
  names <- as.character(rownames(data.frame(tax_table(phy)))[order(data.frame(tax_table(phy))$Subtype)])
  samples <- colnames(typerelabund)
  shores <- data.frame(sample_data(phy))[samples, "Shore"]
  sites <- data.frame(sample_data(phy))[samples, "Site"]
  divides <- apply(typerelabund, 2, cumsum)
  heights <- diff(divides)
  heights[heights < 0.04] <- NA
  # Plot barplot and OTU names and blast hits
  bars <- barplot(typerelabund, col=rainbow(ntaxa(phy)), las=1, space=0.1, xaxs="i",
                  xlab="", ylab="", cex.axis=0.75, cex.names=0.75, xaxt="n")
  text(bars, -0.06, labels=samples, xpd=T, cex=0.75)
  text(bars, -0.10, labels=shores, xpd=T, cex=0.75)
  text(bars, -0.14, labels=substr(sites, 1, 2), xpd=T, cex=0.75)
  text(bars[1]/4, c(-0.06, -0.10, -0.14), labels=c("sample:", "shore:", "site:"), xpd=T, cex=0.75, pos=2)
  mtext(side=2, text = "Relative Abundance", line=2)
  for (i in 1:length(bars)) {
    text(rep(bars[i], length(heights[which(!is.na(heights[,i])),i])), 
         divides[which(!is.na(heights[,i])),i] + heights[which(!is.na(heights[,i])),i] / 2, 
         labels=paste(names[which(!is.na(heights[,i]))+1],blasthits[which(!is.na(heights[,i]))+1],sep="\n"),
         cex=0.5)
  }
  mtext(side=3, main, xpd=T, adj=0, line=0.5, font=2)
}
```

## Individual coral species
### *Diploria strigosa*
```{r dstr, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Diploria strigosa
dstr97.f.p <- subset_samples(phy97.f.p, Species=="strigosa")
dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
# Plot custom barplots for Diploria strigosa
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(dstr97.f.p, main="97% OTUs")
otubarplot(dstr100.f.p, main="100% OTUs")
```

### *Porites furcata*
```{r pfur, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Porites furcata
pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
pfur100.f.p <- subset_samples(phy100.f.p, Species=="furcata")
# Plot custom barplots for Porites furcata
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(pfur97.f.p, main="97% OTUs")
otubarplot(pfur100.f.p, main="100% OTUs")
```

### *Orbicella annularis*
```{r oann, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Orbicella annularis
oann97.f.p <- subset_samples(phy97.f.p, Species=="annularis")
oann100.f.p <- subset_samples(phy100.f.p, Species=="annularis")
# Plot custom barplots for Orbicella annularis
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(oann97.f.p, main="97% OTUs")
otubarplot(oann100.f.p, main="100% OTUs")
```

### *Millepora alcicornis*
```{r malc, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Millepora alcicornis
malc97.f.p <- subset_samples(phy97.f.p, Species=="alcicornis")
malc100.f.p <- subset_samples(phy100.f.p, Species=="alcicornis")
# Plot custom barplots for Millepora alcicornis
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(malc97.f.p, main="97% OTUs")
otubarplot(malc100.f.p, main="100% OTUs")
```

### *Siderastrea siderea*
```{r ssid, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Siderastrea siderea
ssid97.f.p <- subset_samples(phy97.f.p, Species=="siderea")
ssid100.f.p <- subset_samples(phy100.f.p, Species=="siderea")
# Plot custom barplots for Siderastrea siderea
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(ssid97.f.p, main="97% OTUs")
otubarplot(ssid100.f.p, main="100% OTUs")
```

### *Favia fragum*
```{r ffra, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Favia fragum
ffra97.f.p <- subset_samples(phy97.f.p, Species=="fragum")
ffra100.f.p <- subset_samples(phy100.f.p, Species=="fragum")
# Plot custom barplots for Favia fragum
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(ffra97.f.p, main="97% OTUs")
otubarplot(ffra100.f.p, main="100% OTUs")
```

### *Siderastrea radians*
```{r srad, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Siderastrea radians
srad97.f.p <- subset_samples(phy97.f.p, Species=="radians")
srad100.f.p <- subset_samples(phy100.f.p, Species=="radians")
# Plot custom barplots for Siderastrea radians
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(srad97.f.p, main="97% OTUs")
otubarplot(srad100.f.p, main="100% OTUs")
```

### *Porites astreoides*
```{r past, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Porites astreoides
past97.f.p <- subset_samples(phy97.f.p, Species=="astreoides")
past100.f.p <- subset_samples(phy100.f.p, Species=="astreoides")
# Plot custom barplots for Porites astreoides
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(past97.f.p, main="97% OTUs")
otubarplot(past100.f.p, main="100% OTUs")
```

### *Dendrogyra cylindrus*
```{r dcyl, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Dendrogyra cylindrus
dcyl97.f.p <- subset_samples(phy97.f.p, Species=="cylindrus")
dcyl100.f.p <- subset_samples(phy100.f.p, Species=="cylindrus")
# Plot custom barplots for Dendrogyra cylindrus
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(dcyl97.f.p, main="97% OTUs")
otubarplot(dcyl100.f.p, main="100% OTUs")
```

### *Montastraea cavernosa*
```{r mcav, cache=T, fig.height=8, fig.width=8}
# Create subsetted phyloseq objects for Montastraea cavernosa
mcav97.f.p <- subset_samples(phy97.f.p, Species=="cavernosa")
mcav100.f.p <- subset_samples(phy100.f.p, Species=="cavernosa")
# Plot custom barplots for Montastraea cavernosa
par(mfrow=c(2,1), mar=c(2.2,3,2,1), lwd=0.1)
otubarplot(mcav97.f.p, main="97% OTUs")
otubarplot(mcav100.f.p, main="100% OTUs")
```

@@
